---
title: 'Prediction of COVID-19 death rate and effect of policy responses in California'
author: "Lin Wang"
output:
  html_document:
    toc: yes
    toc_float: yes
---
<br>

This is my PM566 Final Project, you can access the full project [here](https://github.com/HopeW233/PM-566-final-project)

<br>

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = FALSE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")

library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(ggplot2)
library(DT)
library(lubridate)
```

## **Introduction**

COVID-19 is a highly contagious disease which affected most states and counties in the United
States. The residents living in areas
with high population density may have more chance to get into close contact with others.

Some countries have been very successful in preventing the spread of the disease, some might have been worse hit than others. Differences in government responses could explain part of this. To understand whether the government policy is effective against the outbreak, it is essential to look at the timing and strictness of responses.

This project is trying to evaluate:

1. Association between COVID death rate and population density in United States.

2. Effect of policy responses to COVID pandemic in California.

<br>

## **Methods**

### 1. Read in the Data

The data was obtained from multiple sources:

National daily COVID data: [COVID19Tracking](https://covidtracking.com/data/download)

Covid data per US states: [NYT] <https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv>

2018 State population data: [NYT] <https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv>

Lockdown data: [Aura Vision's Lockdown Tracker](https://covid19-lockdown-tracker.netlify.com/lockdown_dates.csv)

```{r read in data, echo=FALSE}
download.file("https://covidtracking.com/data/download/national-history.csv", destfile = "national-history.csv")
cv_national <- fread("national-history.csv")

# Getting CA covid history data
download.file("https://covidtracking.com/data/download/california-history.csv", destfile = "california-history.csv")
cv_ca <- fread("california-history.csv")

# Getting covid data per county in CA
download.file("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv", destfile = "us-counties.csv", method="libcurl", timeout = 60)
counties <- fread("us-counties.csv")

download.file("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv", destfile = "us-states.csv", method="libcurl", timeout = 60)
states <- fread("us-states.csv")

#
download.file("https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv", destfile = "us_census_2018_population_estimates_states.csv")
pop <- fread("us_census_2018_population_estimates_states.csv")

#
download.file("https://covid19-lockdown-tracker.netlify.com/lockdown_dates.csv", destfile = "lockdown_dates.csv")
lockdown <- fread("lockdown_dates.csv")

#
cv_states <- merge(
  x = states,
  y = pop,
  by.x = "state",
  by.y = "state_name",
  all.x = TRUE, all.y = FALSE
)

lockdown <- lockdown %>%
  filter(Country == "United States")

cv_states <- merge(
  x = cv_states,
  y = lockdown,
  by.x = "state",
  by.y = "Place",
  all.x = TRUE, all.y = FALSE
)

dat <- cv_states %>%
  select(state, date, cases, deaths, population, pop_density, `Start date`, `End date`)
```

`data.table` is used to download and read in large data, EDA checklist was done to check dimensions, headers, footers, variable names and types.

### 2. Create, rename, format variables, check if any missing values

New variables and some outliers were created or corrected using `data.table`, `tidyverse` and `dplyr`, date variable was re-formatted into correct format by `lubridate`, `NA` in death and cases columns were kept because it would not affect our analysis.

```{r, echo=FALSE}
# selecting columns we need
cv_national <- cv_national[, .(date, states, death, deathIncrease, positive, recovered, totalTestResults)] %>%
  rename(newdeaths = deathIncrease, tests = totalTestResults)
cv_national$date <- as.Date(cv_national$date)

#
cv_national <- cv_national %>%
  mutate(CFR = round((death*100/(death+recovered)), 2)) %>%
  mutate(IFR = round((death*100/positive), 2))

cv_national[CFR == Inf, CFR := 0]
cv_national[IFR == Inf, IFR := 0]
cv_national[death > positive, CFR := 0]
# format
dat$date <- as.Date(dat$date)
dat$`Start date` <- as.Date(dat$`Start date`)
dat$`End date` <- as.Date(dat$`End date`)

state_list <- unique(dat$state)
dat$state <- factor(dat$state, levels = state_list)

# Creating new variables
dat <- dat %>%
  mutate(newcases = cases - lag(cases, 1)) %>%
  mutate(newdeaths = deaths - lag(deaths, 1)) %>%
  mutate(death_rate = round((deaths*100/(cases)), 2)) %>%
  mutate(month = substr(dat$date, 6, 7))
  

dat[newcases < 0, newcases := 0]
dat[newdeaths <0, newdeaths := 0]
dat[death_rate == Inf, death_rate := 0]
dat[deaths > cases, death_rate := 0]

dat$per100k =  as.numeric(format(round(dat$cases/(dat$population/100000),1),nsmall=1))
dat$newper100k =  as.numeric(format(round(dat$newcases/(dat$population/100000),1),nsmall=1))
dat$deathsper100k =  as.numeric(format(round(dat$deaths/(dat$population/100000),1),nsmall=1))
dat$newdeathsper100k =  as.numeric(format(round(dat$newdeaths/(dat$population/100000),1),nsmall=1))

# Creating Quarter variable
counties$date <- as.Date(counties$date)

ca_county <- counties %>%
  filter(state == "California") %>%
  mutate(newcases = cases - lag(cases, 1)) %>%
  mutate(newdeaths = deaths - lag(deaths, 1)) %>%
  mutate(death_rate = round((deaths*100/(cases)), 2))
  

ca_county[newcases < 0, newcases := 0]
ca_county[newdeaths <0, newdeaths := 0]
ca_county[death_rate == Inf, death_rate := 0]
ca_county[deaths > cases, death_rate := 0]
```

### 3. Data Visualization

The tools used to visualize data mainly are `ggplot`, `plotly`.

<br>

## **Results** 

```{r, echo=FALSE, eval=FALSE}
table <- dat %>%
  select(state, date, cases, deaths, newcases, newdeaths, death_rate, per100k, deathsper100k) %>%
 arrange(state, desc(date))

datatable(table)
```

### Figure 1: Ongoing trends of COVID 19 CFR and IFR 

```{r, echo=FALSE}
fig1 <- cv_national %>%
  ggplot()+
  geom_line(aes(date, CFR), color = "skyblue", size = 0.8, na.rm = TRUE)+
  geom_line(aes(date, IFR), color = "#69b3a2", size = 0.8, na.rm = TRUE)+
  geom_vline(aes(xintercept=as.numeric(ymd("2020-03-25"))), linetype = "dashed", color = "red", size = 0.5)+
  geom_vline(aes(xintercept=as.numeric(ymd("2020-02-26"))), linetype = "dashed", color = "darkred", size = 0.5)+
  labs(title = "Figure 1: Overview of COVID 19 CFR and IFR in US", y = "CFR/IFR")+
  theme_bw()

ggplotly(fig1)
```

Figure 1 tells the developing trends of COVID 19 CFR and IFR in US. The green line is infection fatality ratio (IFR), which estimates the proportion of deaths among all infected individuals. The blue line is case fatality ratio (CFR), which estimates the proportion of deaths among identified confirmed cases. During this ongoing epidemic, some of the active cases already detected may subsequently die, leading to underestimation of CFR estimated before their death. In order to mitigate the bias due to delays to case resolution during an ongoing outbreak, the CFR is calculated by this formula: No. of deaths from disease/(No. of deaths from disease+No. of recovered from disease) * 100. However, In this case, CFR may be overestimated if people die quicker than they recover.


### Figure 2: Hypothesis Test of Covid Death Rate

```{r echo=FALSE, fig.height=18, fig.width=18}
dat_today <- dat %>%
  filter(date == max(date)) %>%
  mutate(Mortality = deaths / cases) %>% 
  mutate(Std = sqrt((Mortality * (1 - Mortality))/sqrt(cases)))

mean_mortality <- sum(dat_today$deaths) / sum(dat_today$cases)

fig2 <- dat_today %>%
  ggplot(aes(reorder(state, Mortality), ymin = Mortality - 2 * Std, ymax = Mortality + 2 * Std, color = state, `Death Rate` = round(Mortality*100,2)))+
  geom_point(aes(y = Mortality))+
  geom_hline(yintercept = mean_mortality, linetype = "dashed", color = "red")+
  coord_flip()+
  geom_errorbar()+
  labs(x = "State", y = "Death Rate", subtitle = "Figure 2: Hypothesis Test of Covid Death Rate")+
  theme_bw()+
  theme(text = element_text(size = 18), legend.position = "none")

fig2
```

Figure 2 shows death rate for each state, New York, New Jersey, Massachusetts and Connecticut are states with higher death rates. The red line indicates the mean death rate.

### Figure 3: COVID-19 death rate vs. population density

```{r, echo=FALSE}
fig3 <- dat_today %>%
  filter(state!="District of Columbia") %>%
  plot_ly(x = ~pop_density, y = ~death_rate,
          type = "scatter", mode = "markers", color = ~state, 
          size = ~population, sizes = c(5, 70), marker = list(sizemode='diameter', opacity=0.5),
          hoverinfo = "text",
          text = ~paste(paste(state, ":", sep=""), paste("Cases per 100k: ", per100k, sep="") , paste("Deaths per 100k: ", deathsper100k, sep=""), paste("Death Rate: ", death_rate, sep=""), sep = "<br>"))

fig3
```

```{r eval=FALSE, echo=FALSE}
# Death_rate vs. pop_density

filtered_dat_today <- dat_today %>%
  filter(state!="District of Columbia")

lmDeath_rate = lm(death_rate~pop_density, data = filtered_dat_today)
summary(lmDeath_rate)

# Death_rate vs. pop_density adjusted for population, cases, deaths, newcases, newdeaths
lmDeath_rate2 = lm(death_rate~pop_density + deaths + population + cases + newcases + newdeaths, data = filtered_dat_today)
summary(lmDeath_rate2)


plot(filtered_dat_today$pop_density, filtered_dat_today$death_rate, main = "Figure 3: COVID-19 death rate vs. population density", xlab = "population density", ylab = "death rate")
abline(lmDeath_rate2)

library(sjPlot)
tab_model(lmDeath_rate2, show.ci = FALSE, show.stat = TRUE)

plot_model(lmDeath_rate2, type = "pred", terms = c("cases", "pop_density"), title = "Figure 4: Predicted values of death rate vs. cases")
```

Figure 3 presents a potential positive association between population density and death rate. States with larger population but lower population density have lower death rate, and perhaps location and people behavior differentiate this.

### Figure 4: New cases development in California 

```{r, echo=FALSE}
dat_policy <- dat %>%
  filter(state == "California") %>%
  mutate(policy = ifelse(date %in% `Start date`:`End date`, "first order", 
                         ifelse(date %in% ymd("2020-07-13"):ymd("2020-09-22"),"statewide order",
                                ifelse(date >= ymd("2020-11-21"), "limited lockdown", "loose"))))

fig4 <- dat_policy %>%
  ggplot(aes(x = date, y = newcases))+
  annotate(geom = "rect", xmin = ymd(min(dat_policy$date)), xmax = ymd("2020-03-19"), ymin =0, ymax = 20000, fill = "blue", alpha = 0.2)+
  annotate("text", x = ymd("2020-02-15"), y = 7000, label = "loose")+
  annotate(geom = "rect", xmin = ymd("2020-03-19"), xmax = ymd("2020-05-25"), ymin =0, ymax = 20000, fill = "red", alpha = 0.2)+
  annotate("text", x = ymd("2020-04-20"), y = 7000, label = "first order")+
  annotate(geom = "rect", xmin = ymd("2020-05-25"), xmax = ymd("2020-07-13"), ymin =0, ymax = 20000, fill = "blue", alpha = 0.2)+
  annotate("text", x = ymd("2020-06-15"), y = 7000, label = "loose")+
  annotate(geom = "rect", xmin = ymd("2020-07-13"), xmax = ymd("2020-09-22"), ymin =0, ymax = 20000, fill = "red", alpha = 0.2)+
  annotate("text", x = ymd("2020-08-15"), y = 7000, label = "statewide order")+
  annotate(geom = "rect", xmin = ymd("2020-09-22"), xmax = ymd("2020-11-21"), ymin =0, ymax = 20000, fill = "blue", alpha = 0.2)+
  annotate("text", x = ymd("2020-10-20"), y = 7000, label = "loose")+
  annotate(geom = "rect", xmin = ymd("2020-11-21"), xmax = ymd(max(dat_policy$date)), ymin =0, ymax = 20000, fill = "red", alpha = 0.2)+
  annotate("text", x = ymd("2020-11-21"), y = 7000, label = "limited lockdown")+
  geom_line(color = "#69b3a2")+
  geom_vline(aes(xintercept=as.numeric(ymd("2020-06-18"))), linetype = "dashed", color = "orange", size = 0.5, text = "wear a mask")+ ##wear a mask
  theme_classic()+
  ggtitle("COVID new cases development in California")

ggplotly(fig4)

fig5 <- dat_policy %>%
  ggplot()+
  geom_boxplot(aes(group = policy, y = newcases, color = policy, fill = policy, alpha = 0.5))+
  theme_bw()+
  ggtitle("COVID new cases distribution in California under different policy")

ggplotly(fig5)
```

```{r, echo=FALSE, eval=FALSE}
dat_policy %>%
  ggplot(aes(x = policy, y = newcases))+
  stat_summary(fun.data = "mean_sdl")+
  theme_bw()

dat_policy %>%
  ggplot()+
  geom_boxplot(aes(group = policy, y = newcases, color = policy, fill = policy, alpha = 0.5))+
  facet_wrap(~policy, nrow = 2)+
  theme_bw()
```


Figure 4 show new cases development in California. On March 19, California mandate a state wide safe-at-home order, and on May 25, the order ends and many communities are into reopening plans. On June 18, governor Newsom ordered all Californians to wear face coverings in public places. On July 13, statewide closures were announced. All bars and indoor dining at restaurants must close completely. 

<br>

## **Conclusion**

1. COVID death rate is positively associated with population density. 
2. Policy responses have a positive impact on COVID pandemic.

The downside of this project is that CFR is sophisticated to estimate and lockdown data is not up-to-date.

**Suggestion: stay at home and keep practicing social distance and wearing masks!**


